Nengo: A Python tool for building large-scale functional brain models

### 2013-10-04 nengo.tex
Edits from Eric's feedback, round 2

\begin{abstract}
  Neuroscience currently lacks
  a comprehensive theory of
  how the brain implements cognitive processes
  with biological parts.
  The Neural Engineering Framework (NEF)
  proposes one such theory,
  but has not yet gathered
  significant empirical support,
  partly due to the technical
  challenge of building and simulating
  large-scale models with the NEF.
  Nengo is a tool that builds and simulates
  large-scale models with the NEF.
  Nengo is the primary resource
  for teaching how the NEF
  is used to model neural systems,
  and for doing research
  comparing NEF models
  to experimental data.
  A previous version of Nengo
  implemented in Java was used
  to create Spaun,
  the world's largest functional brain model.
  Simulating Spaun
  identified limitations
  in Nengo's ability to
  construct models with simple syntax,
  simulate them quickly,
  and collect data for subsequent analysis.
  This paper describes a new version of Nengo
  implemented in Python that overcomes these limitations.
  It uses simple and extendable syntax,
  simulates a benchmark model on the scale of Spaun
  200 times faster than the previous implementation,
  and has a flexible mechanism
  for collecting simulation results.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide the dimensionality
(i.e., the length of the vector it represents),
and an object that describes
a population of neurons.
For example, \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
describes an ensemble representing a one-dimensional vector
with 50 leaky integrate-and-fire neurons \citep{lapicque1907}
that have a refractory period of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

Ensembles and nodes can be connected together
in several ways.
Like other PyNengo objects,
a connection contains symbolic information;
in this case, a connection tracks
how two objects are connected together.

@@

However, neurons in an ensemble can be directly connected
to neurons in another ensemble
with synaptic connection weights
using the \texttt{NonlinearityConnection}.
Both of these connection types
can be temporally filtered,
and the weights involved in the connection
(decoding weights for \texttt{DecodedConnection},
synaptic connection weights for \texttt{NonlinearityConnection})
can be modified over time with learning rules.

@@

\begin{figure}
\begin{center}
  \begin{minipage}{.435\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
# We use Nengo to generate
#     bias: Bias currents
#     decoders: Decoding weights
#     encoders: Encoding weights
# for populations A and B.
import pynn.nest as pyNN

lif_params = {'tau_refrac': 2.0, 'tau_syn_E':100,
               'tau_syn_I':100}
dt = 0.5
pyNN.setup(timestep=dt, min_delay=dt)
A = pyNN.Population(30, pyNN.IF_cond_exp, lif_params)
B = pyNN.Population(30, pyNN.IF_cond_exp, lif_params)
input_signal = 0.5
inputA = [pyNN.DCSource(amplitude=val) for val
           in A_bias + input_signal * A_encoders]
for i, pulse in enumerate(inputA):
    pulse.inject_into(A[i:i+1])
inputB = [pyNN.DCSource(amplitude=val) for val
           in B_bias]
for i, pulse in enumerate(inputB):
    pulse.inject_into(B[i:i+1])
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.46\textwidth}
    \begin{lstlisting}[basicstyle={\footnotesize\ttfamily}]
weights = []
for i in xrange(30):
    for j in xrange(30):
        weights.append((i, j,
            np.dot(A_decoder[i], B_encoder[j]), 1.0))
connection = pyNN.Projection(
    A, B, pyNN.FromListConnector(weights))
A.record('spikes')
B.record('spikes')
pyNN.run(1000)

# Decoded value of B
B_output = numpy.zeros(1000)
for i in xrange(1000):
    spikes = B[i:i+1].getSpikes()[:,1]
    spikes = (spikes / dt).astype('int')
    B_output[spikes] += B_decoder[i]
# Filter with tau = 200 ms
decay = np.exp(-dt / 200)
B_output[0, :] *= (1 - decay)
for i in xrange(1, 1000):
    B_output[i,:] = decay * B_output[i-1,:]
                     + (1-decay) * B_output[i,:]
    \end{lstlisting}
  \end{minipage}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:pynn} Figure \arabic{figure}.}{
   Implementation of the communication channel in PyNN.
   Note that several quantities are generated from Nengo;
   the functions that generate these quantities in Nengo
   can be decoupled, but are not included here for brevity.}
\end{figure}

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
NEST \citep{eppler2008}, and PCSim \citep{pecevski2009}.
We run each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
indicating that these means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increase.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run with MPI,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

@@

\section*{Acknowledgement}

We would like to thank
everyone who has contributed
to this reimplementation of Nengo
by providing examples,
unit tests, and bugfixes:
Peter Blouw, Brent Komer, and Bryan Tripp.

% Also acknowledge Bryan for making JavaNengo, and edits

>>>>

\begin{abstract}
  Neuroscience currently lacks
  a comprehensive theory of
  how the brain implements cognitive processes
  with biological parts.
  The Neural Engineering Framework (NEF)
  proposes one such theory,
  but has not yet gathered
  significant empirical support,
  partly due to the technical
  challenge of building and simulating
  large-scale models with the NEF.
  Nengo is a tool that builds and simulates
  large-scale models with the NEF;
  it is the primary resource
  for teaching how the NEF
  is used to model neural systems,
  and for doing research
  comparing NEF models
  to experimental data.
  A previous version of Nengo
  implemented in Java was used
  to create Spaun,
  the world's largest functional brain model.
  Simulating Spaun
  identified limitations
  in Nengo's ability to
  construct models with simple syntax,
  simulate them quickly,
  and collect data for subsequent analysis.
  This paper describes a new version of Nengo
  implemented in Python that overcomes these limitations.
  It uses simple and extendable syntax,
  simulates a benchmark model on the scale of Spaun
  200 times faster than the previous implementation,
  and has a flexible mechanism
  for collecting simulation results.

@@

An \texttt{Ensemble} is
a population of neurons
that represents information
in the form of a real-valued vector.
When creating an ensemble,
the user must provide the dimensionality
(i.e., the length of the vector it represents),
and an object that describes
a population of neurons.
For example,
\begin{quote}
  \texttt{Ensemble('A', nengo.LIF(50, tau\_ref=0.002), 1)}
\end{quote}
describes an ensemble representing a one-dimensional vector
with 50 leaky integrate-and-fire neurons \citep{lapicque1907}
that have a refractory period of 2 milliseconds.
The \texttt{nengo.LIF} class defines
the parameters of the LIF neurons symbolically
so that each simulator can compute
the LIF nonlinearity efficiently.
The neuron model used by the ensemble
is changed by passing in a different symbolic neuron object;
however, the simulator used must be aware
of that type of neuron.

@@

Ensembles and nodes can be connected together
in several ways.
A \texttt{Connection} contains symbolic information
about how two objects are connected.
That information either includes
a factored or full connection weight matrix,
or includes enough information
in order to generate weights during simulation.

@@

However, neurons in an ensemble can be directly connected
to neurons in another ensemble
with synaptic connection weights
using the \texttt{NonlinearityConnection}.
Both connection types
can be temporally filtered,
and the weights involved in the connection
(decoding weights for \texttt{DecodedConnection},
synaptic connection weights for \texttt{NonlinearityConnection})
can be modified over time with learning rules.

@@

\begin{figure}
\begin{center}
  \includegraphics[width=\textwidth]{pynn}
\end{center}
 \textbf{\refstepcounter{figure}\label{fig:pynn} Figure \arabic{figure}.}{
   Implementation of the communication channel in PyNN.
   Note that several quantities are generated from Nengo;
   the functions that generate these quantities in Nengo
   can be decoupled, but are not included here for brevity.}
\end{figure}

@@

The communication channel and Lorenz attractor
are small models that demonstrate
the principles of the NEF.
Their small size enables us to write
PyNN scripts that implement roughly
the same functionality
with Brian \citep{goodman2009}, NEURON \citep{hines2009},
NEST \citep{eppler2008}, and PCSim \citep{pecevski2009}.
We run each parameter set five times
on the same machine,
and plot the mean.
In most cases, the coefficient of variation
is well below 0.1, except for two
outliers with coefficients of 0.18 and 0.22,
indicating that these means are robust.
The results, shown in Figure~\ref{fig:benchmarks}A and B,
suggest that all versions of Nengo are significantly
faster than the simulators accessible
through PyNN, especially
as the size of models increase.
This is likely due to Nengo's
use of factorized weight matrices,
rather than storing and computing with
the entire weight matrix
on each timestep.
While NEST and NEURON were not
run on multiple cores using message passing,
the reference simulator of PyNengo
also only uses one CPU core.
The results also suggest that PyNengo's
simulators are faster than JavaNengo's simulator.

@@

\section*{Acknowledgments}

We thank Bryan Tripp
for editing this paper,
contributing to PyNengo
and the PyNengo OpenCL simulator
and for creating the JavaNengo simulator
that has been useful
for the past six years,
and will continue
to be used for many years to come.
We thank Peter Blouw and Brent Komer,
who have contributed
to PyNengo by providing examples,
unit tests, and bugfixes.
